---
title: "Population validation and cohort representativeness: A short data report in ISARIC and OpenSAFELY"
output:
  html_document:
  theme: journal
github_document:
  keep_html: TRUE
# output:
#   pdf_document

---

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}
pre[class] {
  max-height: 200px;
}
```

<style>
  body {
    text-align: justify;
     font-size: 11pt}
</style>

  
```{r setup, include=FALSE}
# Document settings ----
knitr::opts_chunk$set(echo = TRUE)

# Import libraries ----
library(tidyverse)
library(here)
library(kableExtra)
library(knitr)
library(ggplot2)
library(png)
library(readr)
library(htmltools)
library(gt)
library(scales)
library(lubridate)
library(grid)
library(Gmisc)
library(tidyverse)
library(DiagrammeR)
library(glue)


# Import custom user functions ----
source(here("analysis", "lib", "custom_functions.R"))
source(here("analysis", "lib", "utility.R"))

# Output directory ----
output_dir_rmd <- here("reports", "validation")
fs::dir_create(output_dir_rmd)

# Dates ----
start_date = as.Date("2020-02-01")
end_date = as.Date("2023-01-01")

# Import data ----
processed_isaric <- read_rds(here::here("output", "admissions", "processed_isaric.rds"))
processed_sus_methodA <- read_rds(here::here("output", "admissions", "processed_sus_methodA.rds"))
processed_sus_methodB <- read_rds(here::here("output", "admissions", "processed_sus_methodB.rds"))
processed_sus_methodC <- read_rds(here::here("output", "admissions", "processed_sus_methodC.rds"))

# Combine admissions
admissions_joined_SUS <- processed_sus_methodA %>%
  rbind(processed_sus_methodB %>% 
          filter(!patient_id %in% processed_sus_methodA$patient_id &
                   !first_admission_date_sus %in%processed_sus_methodA$first_admission_date_sus)) %>%
  plyr::rbind.fill(processed_sus_methodC %>% 
          filter(!patient_id %in% processed_sus_methodA$patient_id &
                   !first_admission_date_sus %in%processed_sus_methodA$first_admission_date_sus) %>%
          filter(!patient_id %in% processed_sus_methodB$patient_id &
                   !first_admission_date_sus %in%processed_sus_methodB$first_admission_date_sus))

# Filter dates
ISARIC_admission_1 <- processed_isaric %>% filter(first_admission_date_isaric >= start_date,
                                                  first_admission_date_isaric <= end_date)

SUS_admission_1_method_A <- processed_sus_methodA %>% filter(first_admission_date_sus >= start_date,
                                                             first_admission_date_sus <= end_date)
SUS_admission_1_method_B <- processed_sus_methodB %>% filter(first_admission_date_sus >= start_date,
                                                             first_admission_date_sus <= end_date)
SUS_admission_1_method_C <- processed_sus_methodC %>% filter(first_admission_date_sus >= start_date,
                                                             first_admission_date_sus <= end_date)
admissions_SUS <- admissions_joined_SUS %>% filter(first_admission_date_sus >= start_date,
                                                    first_admission_date_sus <= end_date)

```


#### Contents
* [Purpose](#purpose)
* [Results](#results)
  + [Ascertainment of COVID-19 hospitalisations](#ascertainment)
  + [Consistency and completeness of data](#consistency)

  
# Purpose <a name="purpose"></a>
Given ISARIC is a new, rapidly-deployed data collection initiative, data quality and consistency across different sites may not be well-understood and it is therefore important to validate data with other sources. 

Every ISARIC record corresponds to a "true" COVID-19 admission. Our goal is to see if national hospital data, provided via SUS (APCS or ECDS) is capable of identifying those records. Here we assume that ISARIC, as a tool for identifying COVID-19 in-patient admissions, is 100% specific, but not 100% sensitive. In other words, it may not pick up every COVID-19 admission, either because of

- lack of coverage in certain hospitals
- some admissions were missed in participating hospitals
- admission date was badly recorded

The main purpose of this piece of work is to validate the representativeness of patients captured in ISARIC and OpenSAFELY and the completeness and accuracy of patient data across both datasets. This includes evaluating the generalisability of ISARIC participants in comparison with all hospital admitted patients (as identifiable in HES/SUS data through OpenSAFELY) to establish whether they are representative in terms of their demographics, prior history, mortality, and subsequent recovery.


# Results <a name="results"></a>

## Ascertainment of COVID-19 hospitalisations via ISARIC and SUS <a name="ascertainment"></a>
```{r flow chart, include=FALSE}

# Patients with an COVID-19 hospital admission
n_patients_ISARIC <- nrow(processed_isaric)
n_patients_SUS_method_A <- nrow(processed_sus_methodA)
n_patients_SUS_method_B <- nrow(processed_sus_methodB)
n_patients_SUS_method_C <- nrow(processed_sus_methodC)
n_patients_SUS_joined <- nrow(admissions_joined_SUS)

# Patients with an COVID-19 hospital admission between 2020-02-01 and 2023-01-01
n_ISARIC_admission_1 <- nrow(ISARIC_admission_1)
n_SUS_admission_1_method_A <- nrow(SUS_admission_1_method_A)
n_SUS_admission_1_method_B <- nrow(SUS_admission_1_method_B)
n_SUS_admission_1_method_C <- nrow(SUS_admission_1_method_C)
n_SUS_joined_admission_1 <- nrow(admissions_SUS)

# Patients with an COVID-19 hospital admission in ISARIC but not SUS
n_admission_ISARIC_NOT_SUS <- ISARIC_admission_1 %>%
  filter(!(patient_id %in% admissions_SUS$patient_id)) %>%
  nrow()

# Patients with an COVID-19 hospital admission in SUS but not ISARIC
n_admission_SUS_NOT_ISARIC <- admissions_SUS %>%
  filter(!(patient_id %in% ISARIC_admission_1$patient_id)) %>%
  nrow()

# Patients with an COVID-19 hospital admission in ISARIC and SUS
admission_ISARIC_SUS <- inner_join(ISARIC_admission_1, admissions_SUS, by = c("patient_id" = "patient_id"))
n_admission_ISARIC_SUS <- nrow(admission_ISARIC_SUS)

# Patients with an COVID-19 hospital admission in ISARIC and SUS within 5 days
n_admission_ISARIC_SUS_5days <- admission_ISARIC_SUS %>%
  filter(first_admission_date_isaric <= first_admission_date_sus + 5 |
         first_admission_date_isaric >= first_admission_date_sus - 5) %>%
  nrow()

# Patients with an COVID-19 hospital admission in ISARIC and SUS within 2 days
n_admission_ISARIC_SUS_2days <- admission_ISARIC_SUS %>%
  filter(first_admission_date_isaric <= first_admission_date_sus + 2 |
         first_admission_date_isaric >= first_admission_date_sus - 2) %>%
  nrow()

# Patients with an COVID-19 hospital admission in ISARIC and SUS on the same date
n_admission_ISARIC_SUS_same_date <- admission_ISARIC_SUS %>%
  filter(first_admission_date_isaric == first_admission_date_sus) %>%
  nrow()

```

Between `r format(min(ISARIC_admission_1$first_admission_date_isaric), format = "%d-%b-%Y")` and `r format(max(ISARIC_admission_1$first_admission_date_isaric), format = "%d-%b-%Y")`, a total of `r format(n_ISARIC_admission_1, big.mark = ",", scientific = FALSE)` patients registered at a general practice using TPP software in England were admitted to an acute care hospital contributing data to the ISARIC COVID-19 Clinical Database with a proven or high likelihood of infection with SARS-CoV-2 (Figure 1). During the same period, a total of `r format(n_patients_SUS_joined, big.mark = ",", scientific = FALSE)` patients registered at a general practice using TPP software in England had a COVID-19 related hospital admission as identifiable in HES/SUS data through OpenSAFELY via three methods;

- Method A: Unplanned hospital admission with a ICD10 COVID-19 code as a diagnosis,  N = `r n_patients_SUS_method_A`
- Method B: Any hospital admission with a ICD10 COVID-19 code as a diagnosis, N = `r n_patients_SUS_method_B`
- Method C: A&E attendance resulting in admission to hospital, with a COVID code (from the A&E SNOMED discharge diagnosis refset) as the A&E discharge diagnosis, N = `r n_patients_SUS_method_C`.


Of the `r format(n_ISARIC_admission_1, big.mark = ",", scientific = FALSE)` COVID-19 related hospital admissions recorded in the ISARIC database, `r format(n_admission_ISARIC_SUS, big.mark = ",", scientific = FALSE)` (**`r round(n_admission_ISARIC_SUS/n_ISARIC_admission_1*100, digits = 0)`%**) had a COVID-19 related hospital admission in SUS. Of the `r format(n_SUS_joined_admission_1, big.mark = ",", scientific = FALSE)` COVID-19 related hospital admissions recorded in SUS, `r format(n_admission_ISARIC_SUS, big.mark = ",", scientific = FALSE)` (**`r round(n_admission_ISARIC_SUS/n_SUS_joined_admission_1*100, digits = 0)`%**)  had a COVID-19 related hospital admission in the ISARIC database. 

A total of `r format(n_admission_ISARIC_SUS_same_date, big.mark = ",", scientific = FALSE)` patients (**`r round(n_admission_ISARIC_SUS_same_date/n_ISARIC_admission_1*100, digits = 0)`%** of all first ISARIC admissions) had the same date recorded for their first COVID-19 hospital admission in ISARIC and in SUS. This increased to `r format(n_admission_ISARIC_SUS_2days, big.mark = ",", scientific = FALSE)` (**`r round(n_admission_ISARIC_SUS_2days/n_ISARIC_admission_1*100, digits = 0)`%**) and `r format(n_admission_ISARIC_SUS_5days, big.mark = ",", scientific = FALSE)` (**`r round(n_admission_ISARIC_SUS_5days/n_ISARIC_admission_1*100, digits = 0)`%**) when the window was expanded to within 2 and 5 days, respectively. 

<br>

``` {r, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=10, fig.cap = "**Figure 1 Ascertainment of COVID-19 hospitalisations via ISARIC and SUS.** ", fig.topcaption=TRUE}
grViz(
  glue("digraph my_flowchart {{ 
      graph[splines = ortho]
      node [fontname = Helvetica, shape = box, width = 4, height = 1]
      
        # ISARIC nodes
        ISARIC1[fixedsize = true, width = 6, height = 1.5, label = <Patients registered at a general practice using TPP software <br/> in England who were admitted to an acute care hospital <br/> contributing data to the ISARIC COVID-19 Clinical Database <br/> with a proven or high likelihood of infection with SARS-CoV-2 <br/> N = {n_patients_ISARIC}>]
                
        blank1[label = '', width = 0.01, height = 0.01]
        excluded1[width = 2, label = <Excluded <br/> N = {n_patients_ISARIC - n_ISARIC_admission_1}>]
        
        ISARIC1 -> blank1[dir = none];
        blank1 -> excluded1[minlen = 2];
        {{ rank = same; blank1 excluded1 }}
        
        ISARIC2[fixedsize = true, width = 6, height = 1, label = <Patients with an COVID-19 hospital admission <br/> between 1st February 2020 and 1st January 2023 <br/> N = {n_ISARIC_admission_1}>]
        blank1 -> ISARIC2;
        
        blank2[label = '', width = 0.01, height = 0.01]
        excluded2[width = 2, label = <Patients with an COVID-19 hospital <br/> admission in ISARIC but not SUS <br/> N = {n_admission_ISARIC_NOT_SUS}>]
        
        ISARIC2 -> blank2[dir = none];
        blank2 -> excluded2[minlen = 2];
        {{ rank = same; blank2 excluded2 }}
        
        
        # SUS nodes
        SUS1[fixedsize = true, width = 6, height = 1.5, label = <Patients registered at a general practice using TPP software <br/> in England with a COVID-19 related hospital admission as <br/> identifiable in HES/SUS data through OpenSAFELY <br/> N = {n_patients_SUS_joined}>]

        blank3[label = '', width = 0.01, height = 0.01]
        excluded3[width = 2, label = <Excluded <br/> N = {n_patients_SUS_joined - n_SUS_joined_admission_1}>]

        SUS1 -> blank3[dir = none];
        blank3 -> excluded3[minlen = 2];
        {{ rank = same; blank3 excluded3 }}
        
        SUS2[fixedsize = true, width = 6, height = 1, label = <Patients with an COVID-19 hospital admission <br/> between 1st February 2020 and 1st January 2023 <br/> N = {n_SUS_joined_admission_1}>]
        blank3 -> SUS2;

        blank4[label = '', width = 0.01, height = 0.01]
        excluded4[width = 2, label = <Patients with a COVID-19 hospital <br/> admission in SUS but not ISARIC <br/> N = {n_SUS_joined_admission_1 - n_admission_ISARIC_SUS_same_date}>]

        SUS2 -> blank4[dir = none];
        blank4 -> excluded4[minlen = 2];
        {{ rank = same; blank4 excluded4 }}
        
        SUS3[fixedsize = true, width = 7, height = 2.5, label = <Unplanned hospital admission with a ICD10 COVID-19 code as a diagnosis <br/> N = {n_patients_SUS_method_A} <br/> <br/> Any hospital admission with a ICD10 COVID-19 code as a diagnosis <br/> N = {n_patients_SUS_method_B} <br/> <br/> A and E attendance resulting in admission to hospital, with an COVID-19 code <br/> as the A and E discharge diagnosis <br/> N = {n_patients_SUS_method_C}>]
        SUS1 -> SUS3;
        {{ rank = same; SUS1 SUS3 }}
        
        SUS4[fixedsize = true, width = 7, height = 2.5, label = <Unplanned hospital admission with a ICD10 COVID-19 code as a diagnosis <br/> N = {n_SUS_admission_1_method_A} <br/> <br/> Any hospital admission with a ICD10 COVID-19 code as a diagnosis <br/> N = {n_SUS_admission_1_method_B} <br/> <br/> A and E attendance resulting in admission to hospital, with an COVID-19 code <br/> as the A and E discharge diagnosis <br/> N = {n_SUS_admission_1_method_C}>]
        SUS2 -> SUS4;
        {{ rank = same; SUS2 SUS4 }}

        # Matched nodes
        middle[label = '', width = 0.01, height = 0.01]
        blank2 -> middle[dir = none]; blank4 -> middle [dir = none]
        
        matched[fixedsize = true, width = 8, label = <Patients with a COVID-19 hospital admission in ISARIC and SUS <br/> N = {n_admission_ISARIC_SUS}>]
        middle -> matched;
        
        blank5[label = '', width = 0.01, height = 0.01]
        excluded5[width = 2, label = <Excluded <br/> N = {n_admission_ISARIC_SUS - n_admission_ISARIC_SUS_5days}>]

        matched -> blank5[dir = none];
        blank5 -> excluded5[minlen = 2];
        {{ rank = same; blank5 excluded5 }}
        
        matched2[fixedsize = true, width = 8, label = <Patients whose first COVID-19 hospital admission in ISARIC is within 5 days <br/> of their COVID-19 hospital admission in SUS <br/> N = {n_admission_ISARIC_SUS_5days}>]
        blank5 -> matched2;
        
        blank6[label = '', width = 0.01, height = 0.01]
        excluded6[width = 2, label = <Excluded <br/> N = {n_admission_ISARIC_SUS_5days - n_admission_ISARIC_SUS_2days}>]

        matched2 -> blank6[dir = none];
        blank6 -> excluded6[minlen = 2];
        {{ rank = same; blank6 excluded6 }}
        
        matched3[fixedsize = true, width = 8, label = <Patients whose first COVID-19 hospital admission in ISARIC is within 2 days <br/> of their COVID-19 hospital admission in SUS <br/> N = {n_admission_ISARIC_SUS_2days}>]
        blank6 -> matched3;
        
        blank7[label = '', width = 0.01, height = 0.01]
        excluded7[width = 2, label = <Excluded <br/> N = {n_admission_ISARIC_SUS_2days - n_admission_ISARIC_SUS_same_date}>]

        matched3 -> blank7[dir = none];
        blank7 -> excluded7[minlen = 2];
        {{ rank = same; blank7 excluded7 }}
        
        matched4[fixedsize = true, width = 8, label = <Patients whose first COVID-19 hospital admission in ISARIC and SUS was on the same date <br/> N = {n_admission_ISARIC_SUS_same_date}>]
        blank7 -> matched4;
        
     }}")
)
```
